\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-03-03 papers/jss/dpx4Rev.Rnw
Paper and bib file updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@376 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% Refs to check: Jansen 2002, Dutilh 2009

@@

The data considered here have the general form $\vc{O}_{1:T}=
(O_{1}^{1}, \ldots, O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$,
\ldots, $O_{T}^{1}, \ldots, O_{T}^{m})$ for an $m$-variate time series
of length $T$.  In the following, we use $\vc{O}_{t}$ as shorthand for
$O_{t}^{1}, \ldots, O_{t}^{m}$.  As an example, consider a time series
of responses generated by a single participant in a psychological
response time experiment.  The data consists of three variables,
response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accurate
responding.  These variables are measured on 168, 134 and 137
occasions respectively (the first part of this series is plotted in
Figure~\ref{fig:speed}).  These data are more fully described in
\citet{Dutilh2009}.

@@

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
figure~ref{fig:speed} shows the values of this variable.  The main
question in this experiment was what would happen when this reward
variable would change from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2009}, who provide a
complete description of the experiment and the relevant theoretical
background.

@@

The multinomial logistic model allows us to include covariates on the
initial state and transition probabilities.  \citet{Chung2007} discuss
a related latent transition model for repeated measurement data
($T=2$) using logistic regression on the transition parameters; they
rely on Bayesian methods of estimation.  Covariates on the transition
probabilities can be specified using a one-sided formula as in the
following example:
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
    transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod)
@

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled with the Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial()),
    transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod)
@

@@

In the following model, age is included as covariate on class
membership to test whether, with age, children apply more complex
rules in solving balance scale items.  Similarly to the transition
matrix, covariates on the prior probabilities of the latent states (or
classes in this case), are defined by using a one-sided formula
\code{prior=~age}: 
<<>>= 
data(balance)
set.seed(1)
mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=3,
	family=list(multinomial("identity"), multinomial("identity"),
	multinomial("identity"), multinomial("identity")),
	respstart=runif(24), prior=~age, initdata=balance)
fm <- fit(mod) 
fm
@

@@

To see the ordering of parameters use the following:
<<>>=
setpars(mod, value=1:npar(mod))
@

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
<<>>=
setpars(mod, getpars(mod,which="fixed"))
@
When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
<<>>=
trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial()),
	  trstart=trst,inst=c(.999,0.001))
fm1 <- fit(mod)
@

After this, we use the fitted values from this model to constrain the
regression coefficients on the transition matrix (parameters number~6
and~10):
<<>>=
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm1,equal=conpat)
@

@@

The constructor method return an object of class \code{exgaus}, and is
defined as follows:
<<>>=
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))

@@

Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
<<>>=
mod <- makeDepmix(response=rModels,transition=transition,
    prior=inMod,stat=FALSE)
fm <- fit(mod)
@

@@

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Han van der Maas
provided the speed-accuracy data \cite{Dutilh2009} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2002}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.

>>>>

@@

The data considered here have the general form $\vc{O}_{1:T}=
(O_{1}^{1}, \ldots, O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$,
\ldots, $O_{T}^{1}, \ldots, O_{T}^{m})$ for an $m$-variate time series
of length $T$.  In the following, we use $\vc{O}_{t}$ as shorthand for
$O_{t}^{1}, \ldots, O_{t}^{m}$.  As an example, consider a time series
of responses generated by a single participant in a psychological
response time experiment.  The data consists of three variables,
response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accurate
responding.  These variables are measured on 168, 134 and 137
occasions respectively (the first part of this series is plotted in
Figure~\ref{fig:speed}).  These data are more fully described in
\citet{Dutilh2010}.

@@

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
figure~ref{fig:speed} shows the values of this variable.  The main
question in this experiment was what would happen when this reward
variable would change from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2010}, who provide a
complete description of the experiment and the relevant theoretical
background.

@@

The multinomial logistic model allows us to include covariates on the
initial state and transition probabilities.  \citet{Chung2007} discuss
a related latent transition model for repeated measurement data
($T=2$) using logistic regression on the transition parameters; they
rely on Bayesian methods of estimation.  Covariates on the transition
probabilities can be specified using a one-sided formula as in the
following example:
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
    transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled with the Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial()),
    transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

In the following model, age is included as covariate on class
membership to test whether, with age, children apply more complex
rules in solving balance scale items.  Similarly to the transition
matrix, covariates on the prior probabilities of the latent states (or
classes in this case), are defined by using a one-sided formula
\code{prior=~age}: 
<<>>= 
data(balance)
set.seed(1)
mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=3,
	family=list(multinomial("identity"), multinomial("identity"),
	multinomial("identity"), multinomial("identity")),
	respstart=runif(24), prior=~age, initdata=balance)
fm <- fit(mod,verbose=FALSE) 
fm
@

@@

To see the ordering of parameters use the following:
<<term=FALSE>>=
setpars(mod, value=1:npar(mod))
@

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
<<term=FALSE>>=
setpars(mod, getpars(mod,which="fixed"))
@
When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
<<>>=
trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial()),
	  trstart=trst,inst=c(.999,0.001))
fm1 <- fit(mod,verbose=FALSE)
@

After this, we use the fitted values from this model to constrain the
regression coefficients on the transition matrix (parameters number~6
and~10):
<<term=FALSE>>=
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm1,equal=conpat)
@

@@

The constructor method return an object of class \code{exgaus}, and is
defined as follows:
<<term=FALSE>>=
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))

@@

Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
<<>>=
mod <- makeDepmix(response=rModels,transition=transition,
    prior=inMod,stat=FALSE)
fm <- fit(mod,verbose=FALSE)
@

@@

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Han van der Maas
provided the speed-accuracy data \cite{Dutilh2010} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2002}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.
